---
title: "RRPA2: Find Severe Weather Events Types Which Have the Greatest Impacts on Population Health and Economic Consequences"
output: 
  html_document:
    keep_md: true
---

## Synopsis  

In this report we aim to find the severe weather events types which have the greatest impacts on population health and economic consequences across the United States.   

To find such severe weather events types, we obtained the data from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. The events in the database start in the year 1950 and end in November 2011.   
 
From these data, we found that, in terms of fatalities, injuries, or fatalities plus injuries, TORNADO is the most harmful event type with respect to population health. And EXCESSIVE HEAT is the 2nd most harmful event type with respect to population health in terms of fatalities plus injuries across the United States.     

We also found that in terms of property damge, crop damage, or property damge plus crop damage, event type TORNADO, HAIL, and TORNADO have the greatest economic consequences respectively. And FLASH FLOOD has the 2nd greatest economic consequences in terms of property damge plus crop damage across the United States.  

## Data Processing

The U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. The data comes in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. We obtain the data from the course web site: [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47MB]  

There is also some documentation of the database available. We can find how some of the variables are constructed/defined.  

- National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
- National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)  

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

### Reading the Storm Data

We first unzip the [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) using the tool such as [TarTool](http://tartool.codeplex.com/) and save it as the .csv file "repdata_data_StormData.csv" in the current working directory.

```{r cachedChunk1, warning=FALSE, echo=TRUE, cache=TRUE}
# Set to the current working directory
# Assume repdata_data_StormData.csv is in the current working directory
library(knitr)
opts_chunk$set(echo=TRUE, cache=TRUE, results='markup')

# This chunk will cache its results and only re-evaluate when changes are present
# Load the data 
StormData <- read.csv("./repdata_data_StormData.csv")
```

After reading the storm data, we first check the structure of the dataset
```{r}
str(StormData)
```

Then we check the first few rows. There are 902,297 rows and 37 columns in this dataset.
```{r}
dim(StormData)
```

```{r}
head(StormData[, 1:14])
```

We also check a few types of events (there are 985 different types of events in the dataset)
```{r}
events <- unique(StormData$EVTYPE)
length(events)
```

```{r}
events[1:40]
```


## Results

### Data analysis for which types of events are most harmful with respect to population health across the United States  

For the impact to the populaiton health, we will look at the data variables "FATALITIES" and "INJURIES". 

```{r cachedChunk2}
# Aggregate the (FATALITIES + INJURIES) in dataset for the all types of events 
Fatality_Injury <- aggregate( cbind(FATALITIES, INJURIES) ~  EVTYPE, data=StormData, sum)

# Add a new column "Total" which is the sum of FATALITIES and INJURIES
Fatality_Injury$Total <- Fatality_Injury$FATALITIES + Fatality_Injury$INJURIES
```

In order to show the total number of Fatalities, Injuries, and (Fatalities + Injuries) for all types of events, we make a panel plot of graph with types of events as axis, and the total number of Fatalities, Injuries, and (Fatalities + Injuries) as y-axis. Here, we take the log of the total number to adjust for the skew in the data.

```{r, fig.height=10, fig.width=8}
par(mfrow=c(3,1), mar=c(3,2,2,1), oma=c(3,2,0,0))
plot(Fatality_Injury$EVTYPE, log10(Fatality_Injury$FATALITIES+1), type="p",  
     xlab="", ylab="", main="Fatalities", cex.main=1.0)
lines(Fatality_Injury$EVTYPE,log10(Fatality_Injury$FATALITIES+1), type="l", col="blue") 
plot(Fatality_Injury$EVTYPE, log10(Fatality_Injury$INJURIES+1), type="p",  
     xlab="", ylab="", main="Injuries", cex.main=1.0)
lines(Fatality_Injury$EVTYPE,log10(Fatality_Injury$INJURIES+1), type="l", col="blue") 
plot(Fatality_Injury$EVTYPE, log10(Fatality_Injury$Total+1), type="p",  
     xlab="", ylab="", main="Fatalities + Injuries", cex.main=1.0)
lines(Fatality_Injury$EVTYPE,log10(Fatality_Injury$Total+1), type="l", col="blue") 
mtext("Total Number", side=2, outer=TRUE)
mtext("Types of Events", side=1, outer=TRUE)
```

We look at the new dataset Fatality_Injury to find the max number of Fatalities, Injuries, and (Fatalities + Injuries).

- Find the max number of Fatalities in the new dataset. It is 5,633 and the corresponding event type is TORNADO. 

```{r}
# max number of FATALITIES
max(Fatality_Injury$FATALITIES)
```


```{r, results='markup'}
# max impact of the event type
max_impact_event<- Fatality_Injury[Fatality_Injury$FATALITIES==max(Fatality_Injury$FATALITIES), ][1]
print(max_impact_event, type="html")
```

- Find the max number of Injuries in the new dataset. It is 91,346 and the corresponding event type is TORNADO.

```{r}
# max number of INJURIES
max(Fatality_Injury$INJURIES)
```


```{r, results='markup'}
# max impact of the event type
max_impact_event<- Fatality_Injury[Fatality_Injury$INJURIES==max(Fatality_Injury$INJURIES), ][1]
print(max_impact_event, type="html")
```

- Find the max number of (Fatalities + Injuries) in the new dataset. It is 96,979 and the corresponding event type is TORNADO
```{r}
# max number of total number for (FATALITIES + INJURIES)
max(Fatality_Injury$Total)
```


```{r, results='markup'}
# max impact of the event type
max_impact_event<- Fatality_Injury[Fatality_Injury$Total==max(Fatality_Injury$Total), ][1]
print(max_impact_event, type="html")
```

  
- We also look at the 2nd largest number of (FATALITIES + INJURIES) in  the new dataset. It is 8,428 and the corresponding event type is EXCESSIVE HEAT.
```{r}
Fatality_Injury[Fatality_Injury$Total==max(Fatality_Injury$Total), ][4] <- 0
max(Fatality_Injury$Total)
```


```{r, results='markup'}
# max impact of the event type
max_impact_event<- Fatality_Injury[Fatality_Injury$Total==max(Fatality_Injury$Total), ][1]
print(max_impact_event, type="html")
```

From the above data analysis, we conclude that    

- If we only look at the Fatalities, the event type "TORNADO" is the most harmful with respect to population health across the United States, which has the totla number of 5,633   

- If we only look at the Injuries, the event type "TORNADO" is the most harmful with respect to population health across the United States, which has the totla number of 91,346  

- If we look at the Fatalities together with Injuries, the event type "TORNADO" is the most harmful with respect to population health across the United States, which has the totla number of 96,979  

- If we look at the Fatalities together with Injuries again, the event type "EXCESSIVE HEAT" is the 2nd most harmful with respect to population health across the United States, which has the total numner of 8,428  

### Data analysis for which types of events have the greatest economic consequences across the United States

For the impact to the economic consequences, we will look at the data varibales "PROPDMG", "PROPDMGEXP", "CROPDMG", and "CROPDMGEXP".   

For "PROPDMGEXP" and "CROPDMGEXP", there are some data with non-numeric values such as "", "-", "?", "+", etc. We will process these non-numeric values as NA

```{r cachedChunk3, warning=FALSE}
# Process data for "PROPDMG", "PROPDMGEXP", "CROPDMG", and "CROPDMGEXP" with "EVTYPE
new_data <-StormData[, c("PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP", "EVTYPE")]

# Remove the level info and convert PROPDMGEXP to numeric with "NA" 
new_data$PROPDMGEXP <- as.numeric(levels(new_data$PROPDMGEXP))[new_data$PROPDMGEXP]

# Remove the level info and convert CROPDMGEXP to numeric with "NA" 
new_data$CROPDMGEXP <- as.numeric(levels(new_data$CROPDMGEXP))[new_data$CROPDMGEXP]
```

We will use aggregate to compute the total number for variables "PROPDMG", "PROPDMGEXP", "CROPDMG", and "CROPDMGEXP" for each type of event.


```{r cachedChunk4}
# Aggregate the PROPDMG in dataset for the all types of events 
Prop_Dmg <- aggregate( PROPDMG ~ EVTYPE, data=new_data, sum)

# Aggregate the PROPDMGEXP in dataset for the all types of events 
Prop_DmgExp <- aggregate( PROPDMGEXP ~ EVTYPE, data=new_data, sum, na.action = na.omit)

# Aggregate the CROPDMG in dataset for the all types of events 
Crop_Dmg <- aggregate( CROPDMG ~ EVTYPE, data=new_data, sum)

# Aggregate the CROPDMGEXP in dataset for the all types of events 
Crop_DmgExp <- aggregate( CROPDMGEXP ~ EVTYPE, data=new_data, sum, na.action = na.omit)

```

- Look at the max number for variable "PROPDMG" from dataset "Prop_Dmg". It is 3,212,258 and the event type corresponding to the max number is TORNADO. 
```{r}
# Find the max number for Property damage
max(Prop_Dmg$PROPDMG)
```

```{r, results='markup'}
# max impact of the event type
max_impact_event<- Prop_Dmg[Prop_Dmg$PROPDMG==max(Prop_Dmg$PROPDMG), ][1]
print(max_impact_event, type="html")
```

- Look at the max number for variable "CROPDMG" from dataset "Crop_Dmg". It is 579,596.30 and the event type corresponding to the max number is HAIL.
```{r}
# Find the max number for Crop damage
max(Crop_Dmg$CROPDMG)
```

```{r, results='markup'}
# max impact of the event type
max_impact_event<- Crop_Dmg[Crop_Dmg$CROPDMG==max(Crop_Dmg$CROPDMG), ][1]
print(max_impact_event, type="html")
```

- Look at the data for variables "PROPDMGEXP" and "CROPDMGEXP". 
```{r, fig.height=8, fig.width=8}
par(mfrow=c(2,1), mar=c(3,2,2,1), oma=c(3,2,0,0))
barplot(Prop_DmgExp$PROPDMGEXP, names.arg=Prop_DmgExp$EVTYPE, 
        xlab="", ylab="", main="PROPDMGEXP", col="blue", cex.main=0.8)
barplot(Crop_DmgExp$CROPDMGEXP, names.arg=Crop_DmgExp$EVTYPE, 
        xlab="", ylab="", main="CROPDMGEXP", col="blue", cex.main=0.8)
mtext("Total Number", side=2, outer=TRUE)
mtext("Types of Events", side=1, outer=TRUE)
```

From the plot, we find that the total numbers for "PROPDMGEXP" and "CROPDMGEXP" are very small compared to the total number of "PROPDMG" and "CROPDMG". Thus we will ignore the total numbers for "PROPDMGEXP" and "CROPDMGEXP" and only look at data for variables "PROPDMG" and "CROPDMG".   

- Let's look at the number from PROPDMG, CROPDMG, and (PROPDMG + CROPDMG) by plotting the diagrams
```{r cachedchunk5}
# merge the dataset Prop_Dmg and Crop_Dmg
Property_Corp_Damage <- merge(Prop_Dmg, Crop_Dmg, by="EVTYPE")

# add a new column "Total" which is the sum of PROPDMG and CROPDMG
Property_Corp_Damage$Total <- Property_Corp_Damage$PROPDMG + Property_Corp_Damage$CROPDMG

```

```{r, fig.height=10, fig.width=8}
par(mfrow=c(3,1), mar=c(3,2,2,1), oma=c(3,2,0,0))
plot(Property_Corp_Damage$EVTYPE, log10(Property_Corp_Damage$PROPDMG+1), type="p",  
     xlab="", ylab="", main="Propert Damages", cex.main=1.0)
lines(Property_Corp_Damage$EVTYPE,log10(Property_Corp_Damage$PROPDMG+1), type="l", col="blue") 

plot(Property_Corp_Damage$EVTYPE, log10(Property_Corp_Damage$CROPDMG+1), type="p",  
     xlab="", ylab="", main="Crop Damages", cex.main=1.0)
lines(Property_Corp_Damage$EVTYPE,log10(Property_Corp_Damage$CROPDMG+1), type="l", col="blue") 

plot(Property_Corp_Damage$EVTYPE, log10(Property_Corp_Damage$Total+1), type="p",  
     xlab="", ylab="", main="Property and Crop Damages", cex.main=1.0)
lines(Property_Corp_Damage$EVTYPE,log10(Property_Corp_Damage$Total+1), type="l", col="blue") 
mtext("Total Number", side=2, outer=TRUE)
mtext("Types of Events", side=1, outer=TRUE)
```

- Look at the max number for (PROPDMG + CROPDMG) in the new dataset. It is 3,312,277 and the event type corresponding to the max number is TORNADO.
```{r}
# max number of (PROPDMG + CROPDMG)
max(Property_Corp_Damage$Total)
```

```{r, results='markup'}
# max impact of the event type
max_impact_event<- Property_Corp_Damage[Property_Corp_Damage$Total==max(Property_Corp_Damage$Total), ][1]
print(max_impact_event, type="html")
```

- Let's find the 2nd largest number in this new dataset. We find the 2nd largest number is 1,599,325 and corresponding event type is FLASH FLOOD

```{r cachedchunk6}
# reset the max Total =0 in ordere to find the 2nd largest number
Property_Corp_Damage[Property_Corp_Damage$Total==max(Property_Corp_Damage$Total), ][4] <- 0
max(Property_Corp_Damage$Total)
```

```{r, results='markup'}
max_impact_event <- Property_Corp_Damage[Property_Corp_Damage$Total==max(Property_Corp_Damage$Total), ][1] 
print(max_impact_event, type="html")
```

From the above data analysis, we conclude that    

- If we only look at the Property Damage, the type of event "TORNADO" has the greatest economic consequences across the United States, which has the totla number of 3,212,258 

- If we only look at the Crop Damage, the type of event "HAIL" has the greatest economic consequences across the United States, which has the totla number of 579,596.30

- If we look at the sum of Property Damage and Crop Damage, the type of event "TORNADO" has the greatest economic consequences across the United States, which has the totla number of 3,312,277   

- If we look at the sum of Property Damage and Crop Damage again, the type of event "FLASH FLOOD" has the 2nd greatest economic consequences across the United States, which has the totla number of 1,599,325 

